---
title: "Class 9 Halloween Candy Project"
author: "Joshua Khalil (A17784122)"
format: pdf
toc: true
---

## Background
 In this mini project, you will explore FiveThirtyEight's Halloween Candy dataset. 



## Data Import 


Our data set is a CSV file, we first need to get the data from the FiveThirtyEight GitHub repo.

```{r}
candy<- read.csv("candy-data.csv", row.names=1)
head(candy)

```
>Q1.  How many different candy types are in this dataset?

```{r}
nrow(candy)

```


>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity == 1)
```


One of the most interesting variables in the dataset is `winpercent`. For a given candy this value is the percentage of people who prefer this candy over another randomly chosen candy from the dataset (what 538 term a matchup). Higher values indicate a more popular candy.

```{r}
candy["Twix", ]$winpercent
```



## What is your favorite Candy 

>Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

My favorite candy is Almond Joy 
```{r}
candy["Almond Joy", ]$winpercent
```


>Q4. What is the winpercent value for “Kit Kat”?

The winpercent for kitcat is 76.7686
 
```{r}
candy["Kit Kat", ]$winpercent
```


>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

The winpercent for Tootsie Roll Snack Bars is 49.6535.   
.
```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent 
```


There is a useful skim() function in the skimr package that can help give you a quick overview of a given dataset. Let’s install this package and try it on our candy data.

```{r}
library("skimr")
skim(candy)
```


>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes! Winpercent 

>Q7. What do you think a zero and one represent for the candy$chocolate column?

1 means the candy contains chocolate
0 means the candy does NOT contain chocolate


##  Exploratory analysis

A good place to start any exploratory analysis is with a histogram. You can do this most easily with the base R function hist(). Alternatively, you can use ggplot() with geom_hist(). Either works well in this case and (as always) its your choice.


>Q8.  Plot a histogram of winpercent values

Base R
```{r}
hist(candy$winpercent,
     main = "Distribution of Candy Win Percent",
     xlab = "Win Percent",
     breaks = )
```

GGplot 
```{r}
library(ggplot2)

ggplot(candy, aes(x = winpercent)) +
  geom_histogram(bins=5) +
  labs(title = "Distribution of Candy Win Percent",
       x = "Win Percent",
       y = "Count")


```

>Q9. Is the distribution of winpercent values symmetrical?

The distribution is not perfectly symmetrical. It is typically right-skewed, with more candies clustered around the middle/lower values and a smaller number of very popular candies with high winpercent values.

>Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
median(candy$winpercent)
```


>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
aggregate(winpercent ~ chocolate, data = candy, mean)
```


>Q12. Is this difference statistically significant


```{r}
t.test(winpercent ~ chocolate, data = candy)

```


The P value was much smaller than 0.05 meaning it did differ significantly. 
Chocolate candies have a significantly higher average winpercent than fruity candies, and this difference is statistically significant based on a two-sample t-test

## Overall Candy Rankings

Let’s use the base R `order()` function together with `head()` to sort the whole dataset by winpercent. Or if you have been getting into the tidyverse and the dplyr package you can use the `arrange()` function together with `head()` to do the same thing and answer the following questions:


> Q13. What are the five least liked candy types in this set?

Nik L Nip, Boston Baked Beans, Chiclets,
Super Bubble, and	Jawbusters are the least liked candys.
```{r}
head(candy[order(candy$winpercent),], n=5)
```

>Q14. What are the top 5 all-time favorite candy types?

The top five all times are Reese's Peanut Butter cup, Reese's Miniatures, Twix,Kit Kat and Snickers
```{r}
head(candy[order(candy$winpercent, decreasing = TRUE), ], 5)
```

>Q15. Make a first barplot of candy ranking based on winpercent values

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```


>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col() +
  labs(x = "Win Percent", y = "Candy",
       title = "Candy Rankings (sorted by Win Percent)")

my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"


```
Now let's add some color that signifies candy type


```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```
Now, for the first time, using this plot we can answer questions like:

>Q17. What is the worst ranked chocolate candy?

Sixlets  

>Q18. What is the best ranked fruity candy?

Starburts


## 5 Taking a look at pricepercent

What about value for money? What is the best candy for the least money? One way to get at this would be to make a plot of winpercent vs the pricepercent variable. The pricepercent variable records the percentile rank of the candy’s price against all the other candies in the dataset. Lower values are less expensive and higher values are more


```{r}
library(ggrepel)

ggplot(candy) +
  aes(pricepercent, winpercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```


>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese Miniature offers most bang for buck, providing high winpercent and low pricepercent.


>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
expensive <- candy[order(candy$pricepercent, decreasing = TRUE), ]
top5_expensive <- head(expensive, 5)
top5_expensive

```
```{r}
top5_expensive[which.min(top5_expensive$winpercent), ]
```

Nik L Nip is the least popular of the expensive candy. 

## Exploring the correlation structure

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```
>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity  are most negatively correlated 

>Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent are the most positively correlated.


##  Principal Component Analysis

Let’s apply PCA using the prcomp() function to our candy dataset remembering to set the scale=TRUE argument.

```{r}
pca <- prcomp(candy, scale=T)
summary(pca)
```

```{r}
plot(pca$x[, 1], pca$x[, 2], xlab = "PC1", ylab = "PC2")

```


We can change the plotting character and add some color

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)

```
We can make a much nicer plot with the ggplot2 package but it is important to note that ggplot works best when you supply an input data.frame that includes a separate column for each of the aesthetics you would like displayed in your final plot. To accomplish this we make a new data.frame here that contains our PCA results with all the rest of our candy data. We will then use this for making plots below

```{r}
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```


```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
#library(plotly)
#ggplotly(p)

```

Let’s finish by taking a quick look at PCA our loadings. Do these make sense to you? Notice the opposite effects of chocolate and fruity and the similar effects of chocolate and bar (i.e. we already know they are correlated).

>Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

```{r}
ggplot(pca$rotation) +
  aes(x = PC1, y = reorder(rownames(pca$rotation), PC1)) +
  geom_col()
```
Fruity, pluribus, and hard load strongly in the positive direction of PC1. This makes sense because PC1 separates fruity and hard candies from chocolate and bar candies, a relationship that was previously evident in the correlation analysis and earlier exploratory plots.


## Summary

The correlation analysis showed you how candy attributes relate to each other (chocolate and fruity are essentially mutually exclusive, for instance), and this set the stage for PCA. By reducing the original variables down to a few PCs, you created a “map of candy space” where similar candies cluster together. The PCA loadings revealed that the primary axis of variation separates chocolate-based candies from fruity ones - the same pattern we observed in the correlation matrix, now visualized in a single, interpretable plot 


>Q25.  Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

“winning” candy in this dataset appears to be chocolate-based, bar-style, and not fruity, with additional features like caramel, peanut/almond, or wafer components  These candies tend to have high winpercent values, even when their price is not the lowest, suggesting that consumers are willing to prefer chocolate-forward candies overall.

