---
title: "Class 12: Intro to Genome Informatics Lab"
author: "Joshua Khalil (A17784122)"
format: pdf
---

## Section 1 Proportions of G/G in a population


Downloaded CSV file from Ensemble

Here we read the CSV file 
```{r}
mxl <-read.csv("373531.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl)
```


## Section 4 

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.
The first column is sample name, the second
column is genotype and the third column are the expression values.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```



```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```
lets make a box plot 
```{r}
library(ggplot2)
ggplot(expr) + aes(geno, exp, fill=geno) +
  geom_boxplot(notch = T)
```

