---
title: "Class 5- Data Viz with ggplot"
author: "Joshua Khalil (PID:A17784122)"
format: pdf
toc: true
---

## Background 

There are lots of ways to make figures in R. These include so called "base R" graphics (e.g. `plot()`) qnd tons of add on packages like **ggplot2**.

For example here, we make the same plot with both:
```{r}
head(cars)
```

```{r}
plot(cars)
```
First I need to install the package with the command `install.packages()`

> **N.B** We never run an install cmd in quarto code chunk or we will end up re-installing packages many times which is not what we want!

Everytime we want to use one of these "add-on" packages we need to load it up in R with the `library()` function:

```{r}
library(ggplot2)
```

```{r}
ggplot(cars)
```

Every ggplot needs at least three things:

- The **data**, the information you want plotted 
- The **aes**thitics, how the data map to the plot
- The **geom**etry, the type of plot 

```{r}
p <- ggplot(cars) +
  aes(x=speed, y=dist) +
 geom_point() +
geom_smooth(method="lm", se= FALSE) +
  labs(title="Stopping distance of old cars",
       subtitle = "Data from the `cars` object",
       x= "Speed (mph)",
       y= "Distance (ft)")
```


Add a line to better show relationship 

Render it out 
```{r}
p + theme_bw()
```






## Gene Expression Plot 

We can read the input data from ther class website 
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

A first version plot
```{r}
ggplot(genes) +
  aes(Condition1, Condition2 ) +
  geom_point()
```

```{r}
table( genes$State)
```

Version 2 lets color by `State` so we can see the "up" and "down" significant genes compared to all the unchanging genes 

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point()
```


Version 3 plot, let's modify the default colors to our preferences
```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point() + 
  scale_colour_manual( values=c("Red","Grey","yellow") ) +
labs(x="Control (no drug)",
      y="Drug",
    title="Gene Expression Changes upon GLP-1 drug" ) +
  theme_bw()
  
```

## Going Further 

Let's have a look at the famous **gapminder** data set 
```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)

```


```{r}
head(gapminder,3)
```
Version 1 
```{r}
ggplot(gapminder)+
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point(alpha=0.3,)

```

Let's "facet" (i.e. make a separate plot) by continent rather than the big hot mess

```{r}
ggplot(gapminder)+
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point(alpha=0.3,) +
  facet_wrap(~continent)
```





## Custom Plot 

How big is this gapminder dataset?

```{r}
nrow(gapminder)
ncol(gapminder)
```
 
I want to "filter" down to a subset of this data. I will use the **dplyr** package to help me.

First, I need to install it and load it up...
`install.packages("dplyr")` and then `library(dplyr)`
```{r}
library(dplyr)

gapminder_2007 <- filter(gapminder, year ==2007)
head(gapminder_2007)
```

```{r}
filter(gapminder_2007, country=="Ireland")
```

```{r}
filter(gapminder, year==2007, country=="Ireland")
```

```{r}
filter(gapminder, year==1977, country=="United States")
```

> Q. make a plot comparing 1977 and 2007 for all countries 

```{r}
plotss <- filter(gapminder, year %in% c(1977,2007))
ggplot(plotss)+
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point() +
  facet_wrap(~year)
```

