---
title: "Class 18: Pertussis Mini Project"
author: "Joshua Khalil (PID A17784122)"
format: pdf

---

## Background 

Pertussis whooping cough is a common lung infection caused by the bacteria B. Pertussis. It can infect anyone but it is most deadly for infants (under 1 year of age)


## CDC Tracking data
The CDC tracks the number of Pertussis cases: 

```{r, echo=FALSE}
cdc <-data.frame(
                                year = c(1922L,1923L,1924L,1925L,1926L,
                                         1927L,1928L,1929L,1930L,1931L,
                                         1932L,1933L,1934L,1935L,1936L,1937L,
                                         1938L,1939L,1940L,1941L,1942L,
                                         1943L,1944L,1945L,1946L,1947L,1948L,
                                         1949L,1950L,1951L,1952L,1953L,1954L,
                                         1955L,1956L,1957L,1958L,1959L,
                                         1960L,1961L,1962L,1963L,1964L,1965L,
                                         1966L,1967L,1968L,1969L,1970L,
                                         1971L,1972L,1973L,1974L,1975L,1976L,
                                         1977L,1978L,1979L,1980L,1981L,
                                         1982L,1983L,1984L,1985L,1986L,1987L,
                                         1988L,1989L,1990L,1991L,1992L,1993L,
                                         1994L,1995L,1996L,1997L,1998L,
                                         1999L,2000L,2001L,2002L,2003L,2004L,
                                         2005L,2006L,2007L,2008L,2009L,
                                         2010L,2011L,2012L,2013L,2014L,2015L,
                                         2016L,2017L,2018L,2019L,2020L,
                                         2021L,2022L,2023L, 2024L, 2025L),
        cases = c(107473,164191,165418,152003,
                                         202210,181411,161799,197371,166914,
                                         172559,215343,179135,265269,180518,
                                         147237,214652,227319,103188,183866,
                                         222202,191383,191890,109873,133792,
                                         109860,156517,74715,69479,120718,
                                         68687,45030,37129,60886,62786,
                                         31732,28295,32148,40005,14809,11468,
                                         17749,17135,13005,6799,7717,9718,
                                         4810,3285,4249,3036,3287,1759,
                                         2402,1738,1010,2177,2063,1623,1730,
                                         1248,1895,2463,2276,3589,4195,
                                         2823,3450,4157,4570,2719,4083,6586,
                                         4617,5137,7796,6564,7405,7298,
                                         7867,7580,9771,11647,25827,25616,
                                         15632,10454,13278,16858,27550,18719,
                                         48277,28639,32971,20762,17972,
                                         18975,15609,18617,6124,2116,3044,7063,22538,21996)
      )
```






>Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

I want a plot of year vs cases
```{r}
library(ggplot2)
ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
geom_line()
 
```

>Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice? (add annotation lines for the major milestones of wP vaccination roll-out (1946) and the switch to the aP vaccine (1996).)

```{r}
library(ggplot2)
ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
geom_line() +
geom_vline(xintercept=1946, col= "blue", lty=2)  +
geom_vline(xintercept= 1996, col = "red", lty=2)+
geom_vline(xintercept=2020, col="grey", lty=2)
```
I notice that cases are decreasing from the 1946 point from over 100,000 cases, down to below 10,000 cases, after the development of the wP vaccination that significantly decreased cases from before 1946 to after 1946... However as time progresses the number of cases begins to rise. 

>Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccines,  the cases began to rise again up to 20,000 cases. Once 2020 hit, the cases went back down to less than 10,000 cases. 
This could be because of resistance of the bacteria adapting and mutating over time, or it could be because of less people taking vaccinations out of fear. Or the aP vaccine could not be as strong as the old one, the aP vaccine based on the evidence suggests that the effects of the aP vaccination have much less of an effect on pertussis than the original wP vaccination, proving to be less effective. 

##  Exploring CMI-PB data

The CMI-PB Project < https://www.cmi-pb.org/> mission is to provide the scientific community with a comprehensive, high-quality and freely accessible resource of Pertussis booster vaccination.

Basically make available a large data set on the immune response to Pertussis. They use a booster vaccination as a proxy for Pertussis vaccination. 

They make their data available as JSON format API. We can read this into R with the `read_json()` function from **jsonlite** package

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)
head(subject)
```


>Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

>Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

>Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?
Is this representative of the US Population 

No it is not representative of the US population

```{r}
table(subject$race, subject$biological_sex)
```

We can now read more tables from the CMI-PB database 

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen",
                      simplifyVector = TRUE)
ab_titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer",
                      simplifyVector = TRUE)
```

```{r}
head(specimen)
```
```{r}
head(ab_titer)
```


To make sense of all this data we need to "join" the tables together. Only then will you know that a given antibody measurement (from the `ab_titer` table) was collected on certain days (from the `specimen` table) from a certain wP or aP subject (from the `subject` table). 

We can use **dplyr** and the `*join()` family function to do this. 

>Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)
meta<- inner_join(subject,specimen)
head(meta)
```


>Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

Let's do one more join to join `ab_titer` with all our `meta` data. 
```{r}
abdata <- inner_join(ab_titer,meta)
head(abdata)
```

>Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```


```{r}
table(abdata$antigen)
```

Lets focus on IgG isotype

```{r}
igg <-abdata|>
      filter(isotype== "IgG")
head(igg)
```


>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?







> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens

Make a plot of `MFI_normalised` values for all `antigen` values.
```{r}
ggplot(igg)+
  aes(MFI_normalised,antigen) +
  geom_boxplot()
```


The antigens "PT" "FIM2/3" and "FHA" appear to have the widest range of values 

>Q. Is there a difference for these responses between aP and wP individuals? 

```{r}
ggplot(igg)+
  aes(MFI_normalised,antigen, col= infancy_vac) +
  geom_boxplot()
```



```{r}
ggplot(igg)+
  aes(MFI_normalised,antigen) +
  geom_boxplot() +
facet_wrap(~infancy_vac)
```

>Q. Is there a difference with time (i.e. before booster shot vs after booster shot)

```{r}
ggplot(igg)+
  aes(MFI_normalised,antigen, col= infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit)
```



```{r}

ab.PT.21 <- abdata |>
  filter(dataset == "2021_dataset",
        isotype == "IgG",  
        antigen == "PT")


  ggplot(ab.PT.21) +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    
    geom_line(alpha=0.25, linewidth= 0.6)+
    geom_point(alpha= 0.4, size=1.5) +
    
    
    stat_summary(fun=mean,
                 geom="smooth",
                 se= FALSE,
                 linewidth= 2.8,
                 span=0.5,
                 aes(group= infancy_vac, col= infancy_vac))+
    
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
    
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```


( I confirmed the the professor all the questions he wanted us to answer)


